Forward Sensitivity Equations in the Presence of Events

### 2018-11-29 paper.tex
Update the paper. Some bug fixes

% Define document title and author
	\title{Forward Sensitivity Equations in the Presence of Events}
	\author{Daniel Kaschek and Matthew Fidler
	\thanks{}}
	\markboth{}{}
	\maketitle

% Write abstract here
\begin{abstract}
	Forward sensitivity equations are frequently used in optimization. Based on forward sensitivities, gradient and Hessian of the least squares function can be derived which allow to use gradient-based optimization methods. In the presence of events, i.e. sudden changes of the state variables, the differential equations and corresponding sensitivity equations are structurally unchanged. However, the events on states need to be accounted for as events in the sensitivities. Here, we derive these necessary events and present them in a way that helps with the implementation in computational software. Finally, the sensitivity events are illustrated on a typical example from pharmacometrics.
\end{abstract}

@@

The derivation is based on linearization of the ODE around $x(t_e) = x_e$, this is the value of $x(t)$ at the latest timepoint before the event occurs. With this choice, the linearized ODE reads
\begin{align}
    \dot x \doteq A(p, t) x + b(p, t)
    \label{eq:linearized}
\end{align}
with $A(p, t) = \left.\frac{\partial f}{\partial x}\right|_{x_e}(p, t)$ and $b(p, t) = f(x_e, p, t)- \left.\frac{\partial f}{\partial x}\right|_{x_e}(p, t)x_e$. The general solution of eq.~\eqref{eq:linearized} is
\begin{align}
    x(t) = \Phi(t)x_0 + \Phi(t)\int_0^t \Phi^{-1}(\tau)b(\tau)d\tau,\quad t\leq t_e,
    \label{eq:leqte}
\end{align}
where $\Phi(t) = \big(\varphi_1(t), \dots, \varphi_n(t)\big)$ is the matrix of linearly independent solutions $\varphi_i(t)$ of the homogeneous part of eq.~\eqref{eq:linearized}, i.e., $\forall i:\;\dot\varphi_i = A(p, t)\varphi_i$ with initial condition $\varphi_i(0) = {\rm e}_i$, the $i$'th unit vector.

@@

\subsection{Sensitivities with respect to $t_e$}
Based on the explicit solutions, eqs.~\eqref{eq:linearized} and \eqref{eq:gte}, we find that
\begin{align}
    \left.\frac{\partial x}{\partial t_e}\right|_{t = t_e} =
    \left\{
    \begin{array}{ll}
    0     & \textrm{, for } t\nearrow t_e \\
    -A v_e + \frac{\partial v_e}{\partial t_e} - b(t_e)     & \textrm{, for }t \searrow t_e, 
    \end{array}
    \right.
    \label{eq:jump}
\end{align}
where we have used that $\Phi(0) = \mathbb I$ is the unit matrix. The sensitivities with respect to $t_e$ jump at time point $t_e$. For $t > t_e$, the sensitivities propagate forward in time according to the sensitivity equations. The jump equation contains the derivative of $v_e$ which depends on the kind of the event.\\

@@

\paragraph{Additive} 
A constant $\Delta x_i$ is added to the value of the $i$'th state, $x_i$. In that case, $v_e$ is set to eq.~\eqref{eq:leqte} evaluated at $t_e$ plus the constant $\Delta x_i$. Therefore, the same argumentation as in the continuated case holds and the sensitivity is the same as eq.~\eqref{eq:continued}. The difference $x_e - v_e$ turns out to be either 0 (for the continued states) or $-\Delta x_k$ (for the affected states $k$, in particular $k = i$).\\

\paragraph{Multiplicative}
The value of the $i$'th state, $x_i$, is multiplied with a constant $\alpha_i$. Eq.~\eqref{eq:leqte} is evaluated at $t_e$ and multiplied by $\alpha$ to get $v_e$. Consequently, the derivative $\frac{\partial v_e}{\partial t_e}$ is computed and plugged into eq.~\eqref{eq:jump}, yielding
\begin{align}
    \lim_{t\searrow t_e}\frac{\partial x_i}{\partial t_e} = \left.\frac{\partial f_i}{\partial x}\right|_{x_e}\cdot (x_e - v_e) - (1-\alpha_i)f_i(x_e).
\end{align}
The difference $x_e - v_e$ is either 0 (for the continued states) or $(1-\alpha_k)x_{e, k}$ (for multiplied states).

@@

% Derive sensitivities with respect to the jump value
We use again the explicit solutions eqs.~\eqref{eq:leqte} and \eqref{eq:gte} and find that
\begin{align}
    \left.\frac{\partial x}{\partial v_e}\right|_{t = t_e} =
    \left\{
    \begin{array}{ll}
    0     & \textrm{, for } t\nearrow t_e \\
    \mathbb I     & \textrm{, for }t \searrow t_e.
    \end{array}
    \right.
    \label{eq:value}
\end{align}
Again, the sensitivities jump. They are forward propagated according to their sensitivity equations. Based on eq.~\eqref{eq:value}, we can construct sensitivities in case of additive events, where $v_e = x_e + \Delta x$ with $\Delta x \in \mathbb R^n$, and multiplicative events, where $v_{e, i} = \alpha_i x_{e, i}$ with $i = 1, \dots, n$. This is:
\begin{align}
    \left.\frac{\partial x}{\partial \Delta x}\right|_{t = t_e} = \mathbb I, \qquad \left.\frac{\partial x}{\partial \alpha}\right|_{t = t_e} = {\rm diag}(x_e).
\end{align}

@@

When working with events in numerical ODE solvers, events are typically defined in the format
\begin{center}
\begin{tabular}{|c|c|c|c|}\hline
     var & time & value & method \\
     \hline
\end{tabular}
\end{center}
where ``var'' denotes the state variable, ``time'' is the time point at which the event occurs, ``value'' is the event value and ``method'' is either replace, add or multiply.

Using the results from Section~\ref{sec:main} we are going to derive the set of events that apply for the sensitivities.

@@

In case there are other events associated with the same event time parameter $\tau$ affecting other states $x_k$, the events in line~4 need to be added. Lines~3 and~4 have method = add because these contributions need to be accumulated. The summation taking place in this way corresponds to the scalar product found in eqs.~\eqref{eq:replace} and \eqref{eq:continued}.

In a similar way, we derive the table for additive events, see Table~\ref{tab:additive}.
\begin{table}[ht]
\caption{Additive event and list of required additional sensitivity events.}
\label{tab:additive}
\begin{center}
\begin{tabular}{|r|l|l|l|l|}\hline
     1 & $x_1$ & $\tau$ & $\delta$ & add \\
     \hline
     2 & $\sfrac{\partial x_1}{\partial p_j}$ & $\tau$ & $0$ & add \\
     3 & $\sfrac{\partial x_1}{\partial \tau}$ & $\tau$ & $-J_{11}\delta$ & add \\
     4 & $\sfrac{\partial x_k}{\partial \tau}$ & $\tau$ & $-J_{k1}\delta$ & add \\
     5 &$\sfrac{\partial x_1}{\partial \delta}$ & $\tau$ & 1 & add \\
     \hline
\end{tabular}
\end{center}
\end{table}
In the table, $\delta$ denotes the value which is added to $x_1$ at time $\tau$. The event in line~2 could be omitted. It is shown in the table to explicitly show that $\frac{\partial x_1}{\partial p_j}$ is not affected. The event in line~4 needs to be added if there are other events associated with the same time parameter $\tau$.

Finally, the table for multiplicative events is derived, see Table~\ref{tab:multi}.
\begin{table}[ht]
\caption{Multiplicative event and list of required additional sensitivity events.}
\label{tab:multi}
\begin{center}
\begin{tabular}{|r|l|l|l|l|}\hline
     1 & $x_1$ & $\tau$ & $\alpha$ & multiply \\
     \hline
     2 & $\sfrac{\partial x_1}{\partial p_j}$ & $\tau$ & $\alpha$ & multiply \\
     3 & $\sfrac{\partial x_1}{\partial \tau}$ & $\tau$ & $(1-\alpha)(J_{11}x_1-f_1)$ & add \\
     4 & $\sfrac{\partial x_k}{\partial \tau}$ & $\tau$ & $(1-\alpha)J_{k1}x_1$ & add \\
     5 &$\sfrac{\partial x_1}{\partial \alpha}$ & $\tau$ & $x_1$ & add \\
     \hline
\end{tabular}
\end{center}
\end{table}
In the table, $\alpha$ denotes the value by which $x_1$ is multiplied at time $\tau$. The event in line~4 needs to be added if there are other events associated with the same time parameter $\tau$.

@@

The simulation outcome is shown in Fig.~\ref{fig:simulation}. 
\begin{figure}[ht]
    \centering
    \includegraphics{example.pdf}
    \caption{Simulation of a PK/PD model. (A) The infusion input, parameterized by tlag, tinf and dose, provokes a response in the effect compartment. (B) The effect sensitivities with respect to the input parameters are shown. Start and end of the infusion are indicated by vertical dashed lines.}
    \label{fig:simulation}
\end{figure}
Upon dosing, the effect state is inhibited and recovers after the dosing input switches back to zero, shown in Fig.~\ref{fig:simulation}A. According to the effect sensitivities, see Fig.~\ref{fig:simulation}B, higher doses lead to stronger inhibition. A longer infusion time will decrease the inhibition during the infusion and increase it afterwards. The same holds for an increased lag time. The impact of the lag time is five times higher than the impact of the infusion time. The reason is that the infusion time is connected to the infusion rate $r_1$, i.e., a longer infusion time means a lower infusion rate to keep the dose constant.

@@

\section{Conclusion}
	This section summarizes the paper.

>>>>

% Define document title and author
	\title{Forward Sensitivity Equations in the Presence of Events}

\author{
    \IEEEauthorblockN{Daniel Kaschek\IEEEauthorrefmark{1} and Matthew Fidler\IEEEauthorrefmark{2}}\\
    \IEEEauthorblockA{\IEEEauthorrefmark{1}IntiQuan GmbH, Basel, Switzerland}
    \IEEEauthorblockA{\IEEEauthorrefmark{2}Novartis, Fort Worth TX, USA}
}

	\maketitle

% Write abstract here
\begin{abstract}
	Forward sensitivity equations are frequently used in optimization. Based on forward sensitivities, the gradient and Hessian of the least squares function can be derived which allow to use gradient-based optimization methods. In the presence of events, i.e. sudden changes of the state variables, the differential equations and corresponding sensitivity equations are structurally unchanged. However, the events on states need to be accounted for as events in the sensitivities. Here, we derive these necessary events and present them in a way that helps with the implementation in computational software. Finally, the sensitivity events are illustrated on a typical example from the field of pharmacometrics.
\end{abstract}

@@

The derivation is based on linearization of the ODE around $x(t_e) = x_e$, this is the value of $x(t)$ at the latest time point before the event occurs. With this choice, the linearized ODE reads
\begin{align}
    \dot x \doteq A(p, t) x + b(p, t)
    \label{eq:linearized}
\end{align}
with 
\begin{align*}
A(p, t) &= \left.\frac{\partial f}{\partial x}\right|_{x_e}(p, t)\\
b(p, t) &= f(x_e, p, t)- \left.\frac{\partial f}{\partial x}\right|_{x_e}(p, t)x_e.
\end{align*}
The general solution of eq.~\eqref{eq:linearized} is
\begin{align}
    x(t) = \Phi(t)x_0 + \Phi(t)\int_0^t \Phi^{-1}(\tau)b(\tau)d\tau,\quad t\leq t_e,
    \label{eq:leqte}
\end{align}
where $\Phi(t) = \big(\varphi_1(t), \dots, \varphi_n(t)\big)$ is the matrix of linearly independent solutions $\varphi_i(t)$ of the homogeneous part of eq.~\eqref{eq:linearized}, i.e., $\forall i:\;\dot\varphi_i = A(p, t)\varphi_i$ with initial condition $\varphi_i(0) = {\rm e}_i$, the $i$'th unit vector.

@@

\subsection{Sensitivities with respect to $t_e$}
Based on the explicit solutions, eqs.~\eqref{eq:leqte} and \eqref{eq:gte}, we find that
\begin{align}
    \left.\frac{\partial x}{\partial t_e}\right|_{t = t_e} =
    \left\{
    \begin{array}{ll}
    0     & \textrm{, for } t\nearrow t_e \\
    -A v_e + \frac{\partial v_e}{\partial t_e} - b(t_e)     & \textrm{, for }t \searrow t_e, 
    \end{array}
    \right.
    \label{eq:jump}
\end{align}
where we have used that $\Phi(0) = \mathbb I$ is the unit matrix. The sensitivities with respect to $t_e$ jump at time point $t_e$. For $t > t_e$, the sensitivities propagate forward in time according to the sensitivity equations. The jump equation contains the derivative of $v_e$ which depends on the kind of the event.\\

@@

\paragraph{Additive} 
A constant $\Delta x_i$ is added to the value of the $i$'th state, $x_i$. In that case, $v_e$ is set to eq.~\eqref{eq:leqte} evaluated at $t_e$ plus the constant $\Delta x_i$. Therefore, the same argumentation as in the continued case holds and the sensitivity is the same as eq.~\eqref{eq:continued}. The difference $x_e - v_e$ turns out to be either 0 (for the continued states) or $-\Delta x_i$ (for the affected state $i$).\\

\paragraph{Multiplicative}
The value of the $i$'th state, $x_i$, is multiplied with a constant $\alpha_i$. Eq.~\eqref{eq:leqte} is evaluated at $t_e$ and multiplied by $\alpha$ to get $v_e$. Consequently, the derivative $\frac{\partial v_e}{\partial t_e}$ is computed and plugged into eq.~\eqref{eq:jump}, yielding
\begin{align}
    \lim_{t\searrow t_e}\frac{\partial x_i}{\partial t_e} = \left.\frac{\partial f_i}{\partial x}\right|_{x_e}\cdot (x_e - v_e) - (1-\alpha_i)f_i(x_e).
\end{align}
The difference $x_e - v_e$ is either 0 (for the continued states) or $(1-\alpha_i)x_{e, i}$ (for the multiplied state).

@@

% Derive sensitivities with respect to the jump value
We use again the explicit solutions, eqs.~\eqref{eq:leqte} and \eqref{eq:gte}, and find that
\begin{align}
    \left.\frac{\partial x}{\partial v_e}\right|_{t = t_e} =
    \left\{
    \begin{array}{ll}
    0     & \textrm{, for } t\nearrow t_e \\
    \mathbb I     & \textrm{, for }t \searrow t_e.
    \end{array}
    \right.
    \label{eq:value}
\end{align}
Again, the sensitivities jump. They are forward propagated according to their sensitivity equations. Based on eq.~\eqref{eq:value}, we can construct sensitivities in case of additive events, where $v_e = x_e + \Delta x$ with $\Delta x \in \mathbb R^n$, and multiplicative events, where $v_{e, i} = \alpha_i x_{e, i}$ with $i = 1, \dots, n$. This is:
\begin{align}
    \left.\frac{\partial x}{\partial \Delta x}\right|_{t = t_e} = \mathbb I, \qquad \left.\frac{\partial x}{\partial \alpha}\right|_{t = t_e} = {\rm diag}(x_e).
\end{align}

@@

When working with events in numerical ODE solvers, events are typically specified by 
\begin{center}
\begin{tabular}{|c|c|c|c|}\hline
     var & time & value & method \\
     \hline
\end{tabular}
\end{center}
where ``var'' denotes the state variable, ``time'' is the time point at which the event occurs, ``value'' is the event value and ``method'' is either replace, add or multiply.

Using the results from Section~\ref{sec:main}, we are going to derive the set of events that apply for the sensitivities.

@@

Lines~3 and~4 have method = add because these contributions need to be accumulated when several events occur.

In a similar way, we derive the table for additive events, see Table~\ref{tab:additive}.
\begin{table}[ht]
\caption{Additive event and list of required additional sensitivity events.}
\label{tab:additive}
\begin{center}
\begin{tabular}{|r|l|l|l|l|}\hline
     1 & $x_1$ & $\tau$ & $\delta$ & add \\
     \hline
     2 & $\sfrac{\partial x_1}{\partial p_j}$ & $\tau$ & $0$ & add \\
     3 & $\sfrac{\partial x_1}{\partial \tau}$ & $\tau$ & $-J_{11}\delta$ & add \\
     4 & $\sfrac{\partial x_k}{\partial \tau}$ & $\tau$ & $-J_{k1}\delta$ & add \\
     5 &$\sfrac{\partial x_1}{\partial \delta}$ & $\tau$ & 1 & add \\
     \hline
\end{tabular}
\end{center}
\end{table}
In the table, $\delta$ denotes the value which is added to $x_1$ at time $\tau$. The event in line~2 could be omitted. It is shown in the table to explicitly show that $\frac{\partial x_1}{\partial p_j}$ is not affected.

Finally, the table for multiplicative events is derived, see Table~\ref{tab:multi}.
\begin{table}[ht]
\caption{Multiplicative event and list of required additional sensitivity events.}
\label{tab:multi}
\begin{center}
\begin{tabular}{|r|l|l|l|l|}\hline
     1 & $x_1$ & $\tau$ & $\alpha$ & multiply \\
     \hline
     2 & $\sfrac{\partial x_1}{\partial p_j}$ & $\tau$ & $\alpha$ & multiply \\
     3 & $\sfrac{\partial x_1}{\partial \tau}$ & $\tau$ & $(1-\alpha)(J_{11}x_1-f_1)$ & add \\
     4 & $\sfrac{\partial x_k}{\partial \tau}$ & $\tau$ & $(1-\alpha)J_{k1}x_1$ & add \\
     5 &$\sfrac{\partial x_1}{\partial \alpha}$ & $\tau$ & $x_1$ & add \\
     \hline
\end{tabular}
\end{center}
\end{table}
In the table, $\alpha$ denotes the value by which $x_1$ is multiplied at time $\tau$.

When implementing the additional events, one should keep in mind that:
\begin{enumerate}
	\item States should be changed by the event after changing the sensitivities. Otherwise expressions like $x_1 - \xi$ can accidentally vanish.
	\item Events occurring at the same time point should be implemented as if the occurred at different time points, i.e., changing sensitivities and states in alteration.
\end{enumerate}

@@

The simulation outcome is shown in Fig.~\ref{fig:simulation}. 
\begin{figure}[ht]
    \centering
    \includegraphics{Code/example.pdf}
    \caption{Simulation of a PK/PD model. (A) The infusion input (red dashed line), parameterized by tlag, tinf and dose, provokes a response in the effect compartment (black continuous line). (B) The effect sensitivities with respect to the input parameters are shown. Start and end of the infusion are indicated by vertical dashed lines.}
    \label{fig:simulation}
\end{figure}
Upon dosing, the effect state is inhibited and recovers after the dosing input switches back to zero, shown in Fig.~\ref{fig:simulation}A. According to the effect sensitivities, see Fig.~\ref{fig:simulation}B, higher doses lead to stronger inhibition. A longer infusion time will decrease the inhibition during the infusion and increase it afterwards. The same holds for an increased lag time. The impact of the lag time is five times higher than the impact of the infusion time. The reason is that the infusion time is connected to the infusion rate $r_1$, i.e., a longer infusion time means a lower infusion rate to keep the dose constant.

@@

\section{Conclusion}
The paper used linearization of ODE models to derive left- and right-sided limits of model sensitivities in the presence of different types of events: replacement, addition or multiplication of current state values with an event value at the event time. The expressions found for the jumps of the model sensitivities were interpreted as additional events on the sensitivity states themselves. Thereby, for each event type, a table of additional events to be applied to the sensitivity equations was derived. This makes the entire approach very transparent and easy to implement.
